\section*{Significance Statement}

Much of observational science follows a common recipe: measure two quantities of interest, statistically adjust for confounders, and interpret the remaining correlation as evidence of a causal link. This logic assumes that adjusted correlations are informative about causation. We test this assumption across nine large datasets from health, neuroscience, psychology, genomics, politics, and vision, and find that nearly every pair of measured features is significantly correlated---the so-called ``crud'' background. The pattern is not an artifact of multiple testing: shared variance in real-world data spreads across many dimensions, so standard low-rank adjustments leave a substantial residual correlation distribution. Whether that residual reflects genuine latent dependence, incomplete adjustment, or both is left open by the diagnostic; what it does show is that many small associations reported as significant are not distinctive against same-pipeline background pairs, and within our model (fixed adjustment dimension $K$) increasing the sample size does not resolve the problem. We propose a practical diagnostic---a crud-aware test---that benchmarks any reported correlation against the domain's empirical null. Only associations that clearly exceed this background should be treated as evidence for a direct causal relationship.

\begin{abstract}
Causal claims in observational research typically start with two measured variables (say education and income), adjust for confounders (say parental socioeconomic status), and interpret the remaining correlation as evidence of a causal effect. For this logic to work, adjusted correlations must be informative about their underlying causal effects. We test this assumption using nine large datasets spanning health, neuroscience, psychology, genomics, politics, and vision. Across these datasets, most pairwise correlations are classically significant---the so-called ``crud'' background. We trace this background to approximately $1/f$ eigenvalue spectra, which distribute shared variance across many directions so that the residual crud scale $\sigma_K$ shrinks only slowly with the number of removed components: within the fixed-$K$ regime we model, the moderate-dimensional adjustments we tested (PC regression and Lasso) do not reduce the residual crud scale to negligible levels in any of the nine datasets, regardless of sample size. A decision-theoretic result formalizes the consequence: when the causal signal is comparable to $\sigma_K$, no rule that compresses each pair to a one-dimensional adjusted-association statistic can reliably separate signal from noise. We propose a crud-aware $z$-test that benchmarks observed correlations against the domain's empirical null---the distribution of associations among random feature pairs; associations that are highly significant by classical standards can become entirely unremarkable once background dependence is accounted for. In most domains studied here, even a correlation as large as $r = 0.10$ does not clear the crud-aware threshold.
\end{abstract}
