\subsection{Crud-aware testing}

\paragraph{Normative implication.} The preceding results establish that, in the fixed-$K$ regime we model, the residual crud scale is not driven to zero by moderate-dimensional adjustment, and that when the causal signal is comparable to the crud scale, no rule that compresses each pair to a one-dimensional adjusted-association statistic can reliably separate signal from noise. The practical question is what to do about it. The minimum reporting standard for small-effect observational claims should be to compare any adjusted association to the background distribution produced by the same preprocessing and adjustment pipeline. Standard iid-based correlation tests ask whether an association differs from zero, not whether it is unusual relative to the domain's background correlations. In large datasets this produces a predictable failure mode: many associations have tiny $p$-values even when their magnitudes are typical for the domain. The relevant quantity is therefore a crud-aware percentile (or $p$-value) computed against the domain's empirical null. For a target pair $(i,j)$ with an adjusted association statistic (for example, a residual correlation after removing $K$ PCs), we compare its absolute value to the empirical distribution of absolute adjusted associations over uniformly sampled feature pairs from the same dataset, computed under the identical preprocessing and adjustment pipeline. The resulting crud-aware $p$-value is the fraction of random pairs whose absolute adjusted association exceeds the target's; the crud percentile is its complement (estimated by Monte Carlo sampling). The comparison is magnitude-based (two-sided): we rank by absolute value. For a uniformly random target pair processed through the same pipeline, this empirical $p$-value is (approximately) uniform by construction (Appendix~\ref{app:bridge-lemmas}). For non-randomly chosen targets, the same test controls Type-I error under the null that the target pair's true association is drawn from the domain's background distribution. Practically, reporting an adjusted association should include its percentile within the empirical null; associations below a chosen high percentile of that null (e.g., the 95th) should be reported as not distinctive against the domain background, leaving open whether other lines of evidence---design leverage, mechanism, replication---support a causal interpretation. The 95th and 99th are conventional thresholds, not principled ones; the appropriate cutoff depends on prior odds and decision costs.

\paragraph{Distinction from multiple testing.} Unlike multiple-testing corrections, which assume a sharp null of zero effect, the crud-aware test uses an empirical null built from same-pipeline reference pairs whose residual associations are typically nonzero. It controls Type-I error under the empirical null---the hypothesis that the target pair's association is typical of the background---rather than under the classical null of zero effect. The approach is related to Efron's empirical null framework \citep{efron2004}, which estimates the null distribution of test statistics from data rather than assuming the theoretical $\mathcal{N}(0,1)$. The key difference is that Efron's empirical null corrects for correlation-induced inflation of null z-scores, whereas the crud empirical null reflects residual dependence that survives generic adjustment---whether that dependence stems from genuine latent structure or from incomplete adjustment is left open by the test, which only asks whether the target association is distinctive relative to the same-pipeline background.

\paragraph{Stratified comparisons.} When dependence is heterogeneous, sample random pairs within the relevant comparison class (same modality, ROI, gene family, measurement type) so the baseline matches the target claim, and report which stratum was used.

\subsection{A parametric crud-aware test for small studies}

The empirical null above requires a data matrix large enough to estimate the background distribution of pairwise associations.
Many studies, however, measure only a handful of variables on a small sample---a psychologist recruiting $n=20$ participants and computing a single correlation, for instance.
Such a study can be understood as observing a small submatrix of the much larger data matrix one \emph{could} have collected had the budget permitted measuring all variables in the domain.
The crud factor operates at the level of the full domain: if the typical background correlation among personality measures is $\sigma_{\mathrm{crud}}\approx 0.15$, that background does not disappear simply because the investigator chose to measure only two of them.

The observed sample correlation $r$ differs from the true population correlation $\rho$ because of sampling noise, and $\rho$ itself differs from zero because of background dependence. These are two independent sources of variability, and both contribute to the spread of $r$. Under a Gaussian approximation (Theorem~\ref{thm:parametric-crud} in Appendix~\ref{app:bridge-lemmas}), the total variance of $r$ for a random pair from the domain is the sum of these two contributions: $\mathrm{Var}(r) \approx \sigma_{\mathrm{crud}}^2 + 1/(n-1)$, where $\sigma_{\mathrm{crud}}^2$ is the variance of true correlations across random pairs (the crud) and $1/(n-1)$ is the sampling variance for each pair.
This yields a direct test statistic: for an observed correlation $r$ with sample size $n$ and domain crud level $\sigma_{\mathrm{crud}}$,
\[
z_{\mathrm{crud}} = \frac{|r|}{\sqrt{\sigma_{\mathrm{crud}}^2 + \frac{1}{n-1}}},
\]
with two-sided p-value $p_{\mathrm{crud}} = 2\,\Ph(-z_{\mathrm{crud}})$.
The classical correlation test uses $z_{\mathrm{classical}}=|r|\sqrt{n-1}$, which is the special case $\sigma_{\mathrm{crud}}=0$: it tests against the null of zero population correlation, ignoring that a nonzero correlation may be entirely typical of the domain background.

When $n$ is large, the sampling variance $1/(n-1)$ shrinks but the crud variance does not.
These statements are for fixed adjustment ($K$ fixed), which is the realistic regime when confounder adjustment is bounded by domain knowledge and estimation stability; moreover, indiscriminate adjustment can introduce collider bias, so ``more $K$'' is not always safer.
A study with $n=200$ and $r=0.15$ yields $p<0.05$ classically, yet $p_{\mathrm{crud}}\approx 0.37$ when $\sigma_{\mathrm{crud}}=0.15$: the observed association is entirely typical of the domain background.
In this regime, the classical test is detecting crud, not signal.

\paragraph{Diminishing returns beyond $n \approx 1/\sigma_{\mathrm{crud}}^2$.}
The crud-aware framework also yields a concrete insight about sample size.
The $z$-test denominator $\sqrt{\sigma_{\mathrm{crud}}^2 + 1/(n-1)}$ reveals a natural crossover.
When $n$ is small (specifically $n \ll 1/\sigma_{\mathrm{crud}}^2$), sampling noise is the dominant source of uncertainty, so larger samples genuinely increase power to detect effects. But once $n$ is large enough that the sampling noise $1/(n-1)$ becomes small relative to the crud variance $\sigma_{\mathrm{crud}}^2$, the denominator is approximately $\sigma_{\mathrm{crud}}$ regardless of $n$: the bottleneck is no longer imprecise estimation but the fact that background correlations are themselves nonzero. Additional observations measure the crud more precisely but do not help separate signal from background.
Because $\sigma_{\mathrm{crud}} \approx 0.02$--$0.14$ across the domains in Table~\ref{tab:crud-scale} (at $K=10$), the crossover occurs at roughly $n \approx 50$--$2{,}500$.
Beyond this point, classical $p$-values continue to shrink---explaining why massive studies routinely report ``significant'' but tiny effects---while the crud-aware $p$-value plateaus.
For this specific purpose, there are sharply diminishing returns to collecting more than a few thousand observations beyond the crossover; this does not apply to the many other reasons a study might collect more data---improving precision on a known effect, supporting subgroup analysis, replicating a published finding, or detecting effects that are themselves much larger than the crud scale.
Although larger $n$ can in principle support larger $K$, in high-dimensional data $K$ grows slowly relative to $p$ and estimation noise in adjustment accumulates, so $\sigma_K$ shrinks too slowly to close the gap.

\begin{center}
\fbox{\parbox{0.92\textwidth}{
\textbf{How to use the crud-aware $z$-test.}
\begin{enumerate}
\item Obtain the sample correlation $r$ and sample size $n$ from the study.
\item Look up a domain-appropriate crud scale $\sigma_{\mathrm{crud}}$: use the residual correlation SD from a large reference dataset in your field (Table~\ref{tab:crud-scale} reports values for various domains at $K=10$), or consult published values.
\item Compute $z_{\mathrm{crud}} = |r|/\sqrt{\sigma_{\mathrm{crud}}^2 + 1/(n-1)}$ and compare to standard normal critical values ($z>1.96$ for two-sided $p<0.05$). Report the crud-aware $p$-value alongside the classical one.
\end{enumerate}
We recommend reporting conclusions for a range of plausible $\sigma_{\mathrm{crud}}$ values as a sensitivity analysis.
When a large reference dataset is available, the full empirical null is preferable.
An online calculator implementing this test is at \url{https://koerding.github.io/crud/test.html}.
}}
\end{center}

\subsection{Worked examples}
\label{sec:worked-examples}

The two examples below are illustrative endpoints---a psychometric within-construct pair (designed to load on the same latent trait) and a borderline-significant epidemiological finding---chosen to bracket the magnitude range, not as random draws from the literature.

\paragraph{A large effect that clears the crud scale.} In the HEXACO personality inventory ($p=242$ items, $n=100{,}000$), items within the same personality facet are designed to measure the same underlying trait. The median within-facet correlation is $|r|=0.32$ at $K=0$, and 55\% of within-facet pairs fall in the top 5\% of the empirical null ($11\times$ enrichment over chance). Applying the crud-aware $z$-test with $\sigma_{\mathrm{crud}} = 0.144$ (HEXACO at $K=0$) gives $z_{\mathrm{crud}} = 2.2$ for the median pair, $p_{\mathrm{crud}} = 0.03$. After removing 10~PCs, 29\% of within-facet pairs remain in the top 5\% ($6\times$ enrichment). Known-strong associations clear the empirical null.

\paragraph{A small effect that does not.} In \citet{liu2022}, multivariable models report a nominally significant association between folate and mortality in adults with T2D (folate quartile 1 vs 2 HR 1.17, 95\% CI 1.01--1.37) in an NHANES-derived cohort of adults with type~2 diabetes ($n \approx 8{,}000$). Converting from the reported confidence interval via $z = \log(\mathrm{HR})/\mathrm{SE}(\log\mathrm{HR})$ and $r \approx z/\sqrt{n}$ gives $r \approx 0.02$, giving $p_{\mathrm{classical}} \approx 0.04$. Applying the crud-aware $z$-test with $\sigma_{\mathrm{crud}} = 0.093$ (NHANES at $K=10$) gives $z_{\mathrm{crud}} = 0.24$, $p_{\mathrm{crud}} = 0.81$. Even this barely significant classical result becomes a crud-aware $p$-value of $0.81$: the effect sits squarely in the middle of the empirical null---its magnitude is typical of randomly chosen NHANES variable pairs after the same adjustment. For comparison, known strong biomarker pairs in the same dataset (hemoglobin vs hematocrit, $r=0.91$; ALT vs AST, $r=0.83$) easily clear the crud scale.

Additional worked examples are in SI Appendix~\ref{app:additional-examples}: NHANES biomarkers (positive---known mechanistic pairs clear the crud scale easily) and GTEx RNA-seq (negative---most reported cis-eQTL effects do not).
