\section{Discussion}

When reported adjusted associations are comparable to the crud scale, causal interpretation is fragile without additional design leverage, even if classical $p$-values are tiny. The crud-aware test provides a minimal magnitude standard for deciding when a small association is worth interpreting. As positive controls, smoking and lung cancer (RR$\approx 7$; converted to a Pearson correlation via Eq.~\eqref{eq:rr-to-r} of SI Appendix~\ref{app:scale-bridge} at representative population prevalences $p_X{=}0.20$, $p_Y{=}0.06$, this gives $r\approx 0.28$, or $\approx 3\sigma_K$ against NHANES at $K=10$) and within-facet HEXACO personality items (best pair $>$99.99th percentile) clear the threshold; SI Appendix~\ref{si:smoking-detail} gives the structural-ceiling caveat for binary correlations and the case-control-vs-cohort distinction. Crucially, $3\sigma_K$ alone is not what makes smoking--cancer compelling: decades of converging dose--response, cessation, mechanism, and animal evidence carry that case. The crud-aware test is a magnitude diagnostic, not a sole filter for causal credibility.

\paragraph{Scope and leverage.} Prior knowledge or design-based leverage effectively isolates variables from background dependence. Randomized treatments, discontinuities, policy changes, and valid instruments provide leverage that is orthogonal to generic correlation structure. Informed selection of which pairs to examine---from prior biology, mechanism, or pathway databases---narrows the empirical null; our impossibility result applies to the association statistic itself, not to the prior that selected the pair. Our results do not argue against such approaches; they target association-only workflows that interpret small adjusted pairwise correlations as causal evidence in the absence of additional leverage. The Bayes-error benchmark of Theorem~\ref{thm:assoc-only} (SI Appendix~\ref{app:bridge-lemmas}) bounds magnitude-only classification under a stylized mixture; it is not a claim about inference about $\rho_{ij}$ itself, which improves with $n$, nor about systems that exploit multivariate structure, nonlinear dependence, mechanistic priors, or interventions. The homogeneity of broad spectra and slow $\sigma_K$ shrinkage across our nine datasets is consistent with the obstacle generalizing beyond a single domain or modality; confirming truly cross-domain status requires a broader survey than nine cases can provide.

\paragraph{Empirical comparison to standard alternatives.} On a synthetic data-generating process with latent-factor confounding and a small set of embedded direct-edge pairs (SI Appendix~\ref{si:synthetic-benchmark}), averaged over $50$ seeds, the crud-aware diagnostic at $K=10$ recovers $98\%$ of direct edges at the conventional 95th-percentile rank with precision $\approx 0.05$, and recovers $90\%$ at the 99th percentile with precision $\approx 0.23$ (F1 $\approx 0.36$). This dominates classical $|r|>1.96/\sqrt{n-1}$, BH-FDR (theoretical and dependent-BY variants), and Efron's empirical-null + BH \citep{efron2004} on every metric: BY-FDR and Efron+BH both improve modestly over classical BH (precision $\approx 0.018$ vs.\ $0.015$) but still fail to control nominal FDR because they only recalibrate the marginal distribution of test statistics, not the structural confounding that PC adjustment removes. The benchmark is a single stylized DGP rather than a method-dominance claim; in regimes with sparse rather than broadly-distributed confounding, BH-FDR will outperform. The takeaway is that the crud-aware diagnostic adds a useful magnitude check in the empirical regime the rest of this paper documents, and that it is a meaningful addition over the closest standard competitor (Efron's empirical null) when confounding is structural.

\paragraph{Relation to the causal-inference framework.} The crud-aware test does not replace identification arguments. The standard causal-inference framework---directed acyclic graphs, the do-calculus, instruments, regression discontinuities, sensitivity analyses for unobserved confounding \citep{pearl2009,imbens_rubin2015,cinelli2020,oster2019}---is the primary apparatus for translating observational data into causal claims. A small adjusted correlation can be perfectly informative if the DAG identifies the effect; a large one can be uninformative if confounding is severe. Our test is a magnitude diagnostic complementary to identification: it asks ``is the magnitude distinctive against same-pipeline residual dependence,'' which is a different question from ``is the effect identified.'' The test is most useful when identification is uncertain (the typical observational-claim setting), where it adds a magnitude floor on top of an unverified identification argument; it is least useful when identification is clean (e.g., a well-designed RCT), where magnitude alone is already meaningful relative to a sharp null.

\paragraph{Sensitivity of $\sigma_K$ to discretionary choices.} The reported quantitative multiples ($r/\sigma_K$, etc.) depend on $K$, preprocessing, design type, and reference-dataset choice. The qualitative conclusion---that the residual correlation distribution remains broadly distributed across these choices, and that small adjusted correlations rarely clear the conventional threshold against an approximately matched reference---is robust: $\sigma_K$ at $K=10$ and $K=50$ differ by less than a factor of two in seven of nine datasets (Table~\ref{tab:crud-scale}), and the alt-adjustment results (SI Appendix~\ref{app:alt-adjust}) confirm that Lasso does not collapse the residual distribution either. The quantitative pass/fail at the conventional threshold for a specific paper, however, is sensitive enough that we recommend reporting conclusions across a range of plausible $\sigma_K$ values rather than a single number (SI Appendix~\ref{si:how-to-use}, SI Appendix~\ref{si:reference-catalog}).

\paragraph{Real small effects supported by RCT/mechanism would also fail this test.} Several real, clinically important interventions land at small RR's that the magnitude-only test would not clear: aspirin in primary cardiovascular prevention (pooled $\mathrm{RR}\approx 0.94$ for CV mortality), low-dose statins ($\mathrm{RR}\approx 0.78$ for major CV events), ACE inhibitors in non-diabetic CKD progression. By the magnitude criterion alone, an observational arm of this evidence would land in the bulk of the empirical null. This is not a failure of the test---it is the correct diagnosis: \emph{the magnitude alone is not what makes these effects compelling}. What carries the case in each example is RCT replication, mechanism, dose-response, removal effect on cessation, and biological plausibility---design leverage on top of small magnitudes. A clinician reading the field-level Type-II ratios in SI Appendix~\ref{app:overclaim}, Table~\ref{tab:field-ratios}, should not infer that small-RR clinical effects are spurious; they should infer that small-RR effects \emph{require additional evidence beyond the test}, which is exactly the prescription throughout this paper.

\paragraph{Why the random-pair null is the right reference, even for ``mechanism-driven'' hypotheses.} An applied investigator may object that hypothesizing from mechanism (folate$\to$one-carbon metabolism$\to$methylation, smoking$\to$DNA adducts$\to$lung carcinoma) is not the same as drawing a random pair, so the random-pair null is the wrong reference. In narrow settings the objection has force---a single pair pre-specified before any data was looked at, tested in an independent dataset. In practice, however, ``I had a mechanism'' is often a post-hoc rationalization of one story among many that could have been told. The literature contains thousands of plausible mechanistic narratives; selecting which one to test, in which dataset, with which adjustment set, is itself a form of multiple selection. The applied user typically samples mechanisms from a vast latent pool and reports the ones that survive; the magnitude distribution of the reported subset is much closer to the random-pair magnitude distribution than the user's introspective sense of being mechanistically pre-specified would suggest. The crud-aware percentile rank treats this honestly: it asks whether the reported magnitude is what one would see by drawing pairs at random from the domain. If the investigator's mechanism is genuinely privileged, the answer should distinguish it from the bulk; if the mechanism is one of many such stories, the random-pair null is the appropriate reference. The test does not adjudicate which case the investigator is in---that is a question of design, mechanism, and replication---but the random-pair reference is defensible precisely because the alternative (``my mechanism is privileged, so I deserve a narrower null'') is rarely justified in practice.

\paragraph{What to do when an effect does not clear the test.} Failing the test is not a verdict that the relation is causally unimportant; it is a verdict that the magnitude alone cannot carry the case. Constructive next steps are: (i) seek design leverage---randomization, instruments, regression discontinuities, negative controls---which breaks the generic-dependence problem the test diagnoses; (ii) pre-register the hypothesis and replicate in an independent dataset, which is the most reliable correction for selection-driven inflation; (iii) accept that converging evidence (mechanism, dose--response, removal effect, prior literature) can carry a causal claim even when magnitude alone does not, as in the smoking--cancer case. For fields where (i)~is structurally unavailable---nutritional, environmental, and lifestyle epidemiology, where randomization of long-term exposure is infeasible and instruments and discontinuities are vanishingly rare---the honest fallback is (iv): report the magnitude rank as a descriptive feature of the result (``$r=0.02$, crud-aware percentile $0.19$''), do not claim causation from magnitude alone, and concentrate causal-interpretation language on the converging-evidence chain (Mendelian randomization where applicable, negative controls, dose-response, biological plausibility) rather than on the magnitude. When neither design leverage nor a converging-evidence chain is available and the magnitude is in the bulk of the empirical null, the appropriate reporting standard is to label the association as not distinctive against background and treat any causal interpretation as provisional.

\paragraph{Limitations.} All results concern linear dependence (correlations, linear PC adjustment, covariance spectra). In domains with substantial nonlinear dependence---gene regulatory networks, neural population codes, epidemiological dose--response---$\sigma_K$ is a lower bound on the true background dependence; whether nonlinear adjustment would shrink the nonlinear background faster than linear adjustment shrinks the linear background is not tested here. The quantitative $\sigma_K$ varies across domains and preprocessing; we use PC regression as a deliberately generic adjustment, not a claim of causal identification. In high-dimensional regimes ($p$ not small relative to $n$), estimating dominant directions adds uncertainty to $\sigma_K$ that the test does not propagate (SI Appendix~\ref{app:alt-adjust}). The decision-theoretic limit is per-pair, equal-variance, Gaussian, and fixed-$K$. SI Appendix~\ref{si:practical-stakes} discusses asymmetric costs of skepticism in concrete settings (e.g., NIH R01 marginal-grant analysis). Three foundational issues are out of scope here and remain open: an analytic correction for selected target pairs (we recommend pre-registration and stratified references but provide no closed-form fix); whether the empirical null reflects latent dependence or unmodeled adjustment (we proceed under the practical position that magnitude typical of same-pipeline pairs is not by itself causal evidence, regardless of which interpretation prevails); and broader robustness of the impossibility theorem beyond the stylized Gaussian-mixture model.
