\subsection*{Datasets.} Nine large public datasets spanning population health, neuroscience, genomics, psychology, political science, and vision (Section~\ref{sec:datasets-main}; SI Appendix~\ref{si:datasets} gives provenance and dimensions): NHANES, precinct voting + Census, GTEx v8, Allen Institute mouse RNA-seq, HEXACO, CIFAR-10, Kay fMRI, Haxby fMRI, and Stringer mouse V1.

\subsection*{Preprocessing.} Each feature is z-scored (mean zero, unit variance). Sample sizes used in analysis are capped at $n\approx 10{,}000$ per dataset for computational tractability where the raw $n$ is larger.

\subsection*{Adjustment.} The deliberately generic adjustment is regression on the top $K$ principal component scores, with $K=10$ as the headline choice and sensitivity reported across $K\in\{0,1,10,50\}$ (Table~\ref{tab:crud-scale}). PCA is computed via randomized SVD with up to $300$ components. Alternative adjustments (Lasso) are reported in SI Appendix~\ref{app:alt-adjust}.

\subsection*{Empirical null and crud scale.} For each dataset and each $K$, we compute residual correlations across uniformly sampled feature pairs (or, where $\binom{p}{2}\le 5\times 10^5$, all pairs). The crud scale $\sigma_K$ is the SD of these residual correlations. Eigenvalue spectra are fit by power-law on the first 40 components against an exponential alternative via AIC.

\subsection*{Statistical analysis.} The crud-aware $z$-test compares an observed correlation $r$ to $\sigma_K$ via Equation~\eqref{eq:z-crud-main}. The full empirical null replaces the Gaussian formula with the empirical fraction of reference pairs whose absolute adjusted association exceeds the target's. Calibration of the parametric formula against the empirical null in heavy-tailed datasets is reported in SI Appendix~\ref{app:bridge-lemmas}, Table~\ref{tab:heavy-tail-cal}. A reference catalog mapping common application domains to a recommended $\sigma_K$ is in SI Appendix~\ref{si:reference-catalog}.

\subsection*{Code and data.} All analysis is in the Python package at \url{https://github.com/koerding/crud}. An interactive demonstration and a standalone crud-aware test calculator are at \url{https://koerding.github.io/crud/} and \url{https://koerding.github.io/crud/test.html}.
