\section{Evidence that small-effect overclaiming is widespread}
\label{app:overclaim}

The main text argues that writing as if small adjusted associations were substantively meaningful is a widely shared practice across empirical sciences. This appendix gives a compact, field-by-field summary of the meta-research behind that claim. Each row of the table is anchored on a single canonical meta-research paper, and every quantitative claim in the row is taken directly from that paper, so each row can be audited against one source. Power/inflation and effect-size distributions are linked mechanistically: \citet{gelman2014} show that significance filtering under low power inflates published effect magnitudes (Type~M error), and \citet{ioannidis2008winner} documents the corresponding empirical pattern across disciplines. Table rows are ordered from the most observational fields (for which the crud-scale argument applies most directly) to the most design-mixed.

\paragraph{How this table was built.}
Fields were selected to span the empirical sciences where small adjusted observational associations are routinely interpreted as causal evidence. Within each field we anchor the row on a single widely-cited meta-research paper that quantifies either statistical power, effect-size inflation, or the distribution of published effect sizes; all numbers in the row come from that paper. We did not perform a de novo meta-analysis. For ``Design mix'', clean published experimental-vs-observational shares are not available for most fields; entries are a qualitative dominant-design label. All numbers are summary statistics of effect-size distributions, not single-study point estimates, and are intended as order-of-magnitude benchmarks.

{\footnotesize
\renewcommand{\arraystretch}{1.2}
\setlength{\tabcolsep}{3pt}
\begin{longtable}{p{0.11\linewidth} p{0.09\linewidth} p{0.11\linewidth} p{0.18\linewidth} p{0.18\linewidth} p{0.21\linewidth}}
\caption{Field-by-field evidence that small published effects are often underpowered and inflated, paired with the prevailing study-design mix and with the typical effect-size magnitudes that published work actually discusses. Each row is anchored on a single meta-research paper; every number in the row is taken verbatim from that paper, in the units the paper itself uses ($\mathrm{RR}$, $r$, $r^2$, or fraction of genetic variance). Rows are ordered from most observational (top) to most design-mixed (bottom). The conversions between these units (Pearson $|r|$ from $\mathrm{RR}$ or from $r^2$) used in the surrounding discussion are derivations, not numbers from the cited papers. The crud-aware test complements power- and inflation-based diagnostics: even when a study is well-powered for its point null, a statistically significant adjusted association of typical field magnitude may still be indistinguishable from the domain's background dependence.\label{tab:overclaim-evidence}} \\
\toprule
Field & Design mix & Reference & What the paper shows & Key number from that paper & Typical published effect size reported \\
\midrule
\endfirsthead
\multicolumn{6}{c}{\tablename\ \thetable\ -- \textit{continued from previous page}} \\
\toprule
Field & Design mix & Reference & What the paper shows & Key number from that paper & Typical published effect size reported \\
\midrule
\endhead
\midrule
\multicolumn{6}{r}{\textit{continued on next page}} \\
\endfoot
\bottomrule
\endlastfoot
Medicine / observational epidemiology &
Predominantly observational &
\citet{siontis2011} &
Survey of nominally significant risk-factor and intervention associations with tiny effects in the published medical literature. &
51 nominally significant tiny-effect estimates ($p<0.05$, $\mathrm{RR}\in[0.95,1.05]$) reported in 40 articles; in 22 of those 40 articles (55\%) the authors expressed no concern about effect-size magnitude. &
Reported $\mathrm{RR}$ is bounded by construction at $0.95\le\mathrm{RR}\le 1.05$. \\
\midrule
Nutritional epidemiology &
Predominantly observational &
\citet{schoenfeld2013} &
``Cookbook review'' of cancer associations for 50 common ingredients drawn from a random recipe book. &
40 of 50 ingredients had at least one cancer study; among those primary studies, 39\% concluded increased cancer risk, 33\% decreased risk, 5\% borderline, and 23\% no association. &
Across the corresponding meta-analyses, $\mathrm{RR}$s were on average null: median $\mathrm{RR}=0.96$, IQR $0.85$--$1.10$. \\
\midrule
Genetics (common-variant) &
Observational by design &
\citet{park2010} &
Direct estimation of the effect-size distribution for SNPs identified by GWAS, fitted to height, Crohn's disease, and breast/prostate/colorectal (BPC) cancers. &
Per-variant effect sizes of the SNPs detected up to that point span $0.04$--$1.96\%$ of genetic variance across the studied traits. &
For height, $\sim$201 estimated susceptibility loci jointly explain $\sim$16\% of genetic variance; the estimated underlying distribution implies many more loci with even smaller per-variant effects remain undiscovered. \\
\midrule
Neuroscience (BWAS) &
BWAS observational &
\citet{marek2022} &
Brain-wide association studies in three large neuroimaging cohorts; tests the sample size needed for reliable detection. &
Reliable detection of typical brain--behavior associations requires sample sizes in the thousands. &
Median univariate brain--behavior $|r|\approx 0.01$; top $1\%$ of associations reach $|r|>0.06$; single largest replicated effect $|r|=0.16$. \\
\midrule
Psychology &
Mixed (experiment-leaning) &
\citet{richard2003} &
Century-scale quantitative synthesis of social-psychology effect sizes. &
474 meta-analytic effects synthesized from $>$25{,}000 primary studies. &
Mean $|r|\approx 0.21$ across all synthesized effects. \\
\midrule
Economics &
Predominantly observational &
\citet{ioannidis2017econ} &
Power meta-analysis of the empirical economics literature, surveying 159 areas of research drawing on 64{,}076 estimates from over 6{,}700 studies. &
Median statistical power is 18\% or less; nearly 80\% of adequately powered reported effects are exaggerated, typically by a factor of two. &
Average effect size across the surveyed estimates (measured largely as partial correlations and elasticities) is $0.17$. \\
\midrule
Behavioural ecology / evolution &
Mixed (lab + field) &
\citet{mollerjennions2002} &
Synthesis of effect sizes across ecology and evolution meta-analyses. &
43 published meta-analyses synthesized, comprising 93 mean-effect estimates using Pearson's $r$ and 136 using Hedges' $d$ or $g$. &
Mean variance explained $r^2 = 2.51$--$5.42\%$ across the synthesized estimates (95\% CIs $1.99$--$7.05\%$). \\
\end{longtable}
}

A consistent pattern emerges. The reported magnitudes are in different native units across rows, but standard conversions place them on a common scale. Pearson $|r|$ values are reported directly only by \citet{marek2022} (median brain--behavior $|r|\approx 0.01$, single largest replicated $|r|=0.16$), \citet{richard2003} (mean $|r|\approx 0.21$), and \citet{ioannidis2017econ} (average partial correlation/elasticity $\approx 0.17$). For the remaining rows, applying the binary RR-to-$r$ conversion of Equation~\eqref{eq:rr-to-r} (Appendix~\ref{app:scale-bridge}) and the $|r|=\sqrt{r^2}$ identity gives: $\mathrm{RR}\in[0.95,1.05]\Rightarrow |r|\lesssim 0.025$ (medicine, evaluated at high outcome prevalence; at the low prevalences typical of rare-disease epidemiology, $|r|$ is several-fold smaller), median $\mathrm{RR}=0.96\Rightarrow |r|\approx 0.02$ (nutrition, similarly prevalence-sensitive), per-variant variance $\le 1.96\%$ of genetic variance combined with heritabilities of order $0.5$--$0.8$ implying per-variant $|r|\lesssim 0.13$ (genetics), and $r^2=2.5$--$5.4\%\Rightarrow |r|\approx 0.16$--$0.23$ directly (behavioural ecology). All converted values inherit the caveats in Appendix~\ref{app:scale-bridge}. Together the typical published-effect magnitudes across the seven fields cluster in the range $|r|\approx 0.01$--$0.23$. Critically, these typical effect sizes are of the same order as---and in several fields smaller than---the empirical crud scales we report across nine datasets (Table~\ref{tab:crud-scale}, where $\sigma_{\mathrm{crud}}^{(K=0)}$ ranges from $0.04$ to $0.19$). The ``Design mix'' column also locates where the paper's contribution is sharpest: in the four fields that are essentially entirely observational (medicine / epidemiology, nutritional epidemiology, common-variant genetics, BWAS-style neuroscience), the crud-scale argument applies directly; in the three fields with substantial experimental or quasi-experimental components (psychology, economics, behavioural ecology), randomization or design-based identification breaks the generic-dependence problem for the experimental subset, but power and inflation issues persist. The interpretive step this paper targets---treating a small, statistically significant adjusted association as support for a causal claim---is therefore being taken under conditions where, on average, the cited meta-research finds power inadequate and reported effects inflated, and against a background of generic dependence that is numerically of similar magnitude. Our contribution is orthogonal and complementary to the power/inflation literature: we ask, in addition, whether a given small adjusted association is distinguishable from the domain's empirical null.

\paragraph{Would the typical published effect clear the crud-aware threshold?} A direct way to read the table above against the test we propose is to ask, for each field, whether the typical reported magnitude exceeds the conventional 95th-percentile crud-aware threshold ($z_{\mathrm{crud}}>1.96$, i.e.\ $|r|>2\sigma_K$) against an approximately matched reference $\sigma_K$ from Table~\ref{tab:crud-scale} at $K=10$. The matchings below are illustrative, not domain-specific---the right comparison for a given paper would use a same-pipeline same-population $\sigma_K$.

\begin{table}[h]
\centering
\caption{Field-level signal-to-crud ratios. ``Typical $|r|$'' is taken from the converted values in the conclusion above; reference $\sigma_K$ is matched approximately to each field's data modality at $K=10$ (Table~\ref{tab:crud-scale}); ``Verdict'' is whether the typical effect clears $|r|/\sigma_K>2$. Caveats from Appendix~\ref{app:scale-bridge} apply throughout. The verdict is a verdict on \emph{magnitude}, not on causation: small-RR effects supported by RCT or mechanism (e.g., aspirin RR$\approx 0.94$ for CV mortality, statins RR$\approx 0.78$ for major CV events) would also fail this magnitude test, and rightly so---the test asks whether magnitude alone separates a finding from typical residual dependence, not whether the underlying effect is real.}
\label{tab:field-ratios}
\begin{tabular}{lcccl}
\toprule
Field & Typical $|r|$ & Reference $\sigma_K$ (dataset) & $|r|/\sigma_K$ & Verdict \\
\midrule
Medicine (RR$\sim$1.05) & $0.025$ & $0.093$ (NHANES) & $0.27$ & fails \\
Nutrition (median RR$\,0.96$) & $0.04$ & $0.093$ (NHANES) & $0.43$ & fails \\
Genetics (per-variant) & $\lesssim 0.13$ & $0.071$ (RNASeq) & $\lesssim 1.8$ & fails / marginal \\
Neuroscience (BWAS median) & $0.01$ & $0.035$ (Kay fMRI) & $0.29$ & fails \\
Economics & $0.17$ & $0.093$ (NHANES) & $1.8$ & marginal \\
Beh.\ ecology & $0.18$--$0.23$ & --- & --- & no analog \\
\bottomrule
\end{tabular}
\end{table}

The pattern is striking: in observational medicine, nutritional epidemiology, common-variant genetics, and BWAS-style neuroscience, the typical published effect would \emph{not} clear the crud-aware threshold against a matched reference. Economics lands near the threshold; behavioural ecology has no clean analog in our nine datasets. Psychology is omitted from the table: the row's natural anchor (Richard 2003's mean $|r|\approx 0.21$) draws from meta-analyses with substantial experimental content, and we lack a meta-analytic synthesis of predominantly observational psychology of comparable scope to anchor the row. The HEXACO worked example in Section~\ref{sec:worked-examples} provides the qualitative picture for the observational individual-differences subset (within-facet item pairs clear $\sigma_K$ comfortably). The fields whose typical effects fall below the threshold are the same fields the meta-research literature has flagged for power and inflation problems---the test is not adding a new diagnosis, but it is providing a quantitative pass/fail at the per-paper level that the power/inflation literature has not.
